---
title: "Block Maxima vs POT for the H\u00e9non Map"
output: html_document
---

This vignette compares block maxima and peaks-over-threshold approaches on data generated from the H\u00e9non map.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Generate H\u00e9non Map Data

```{r}
library(chaoticds)
orbit <- simulate_henon_map(5000, a = 1.4, b = 0.3)
series <- orbit$x
```

## Block Maxima Analysis

```{r}
library(chaoticds)
blocks <- block_maxima(series, block_size = 50)
gev_fit <- tryCatch(fit_gev(blocks), error = function(e) e)
print(gev_fit)
```

## Peaks Over Threshold Analysis

```{r}
library(chaoticds)
threshold <- quantile(series, 0.95)
exc <- exceedances(series, threshold)
print(length(exc))
```

Both approaches provide complementary views of the tail behaviour. The block
maxima method models the distribution of maxima in fixed blocks, while the
POT approach focuses on exceedances over a high threshold. Comparing the two can
reveal whether a single GEV model adequately describes extremes of the Hénon
map.

