---
title: "Estimating \u03b8 for the Logistic Map"
output: html_document
---

This vignette demonstrates how to estimate the extremal index of the logistic map using the utilities in this repository.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Simulate the Logistic Map

```{r}
library(chaoticds)
series <- simulate_logistic_map(5000, r = 3.8, x0 = 0.2)
```

## Estimate the Extremal Index

```{r}
library(chaoticds)
threshold <- quantile(series, 0.95)
theta_runs <- extremal_index_runs(series, threshold, run_length = 5)
theta_intervals <- extremal_index_intervals(series, threshold)
print(c(runs = theta_runs, intervals = theta_intervals))
```

## Visualize Hitting Times

```{r}
hts <- hitting_times(series, threshold)
plot_hts(hts)
```

The plot compares empirical hitting-time survival to the theoretical exponential
curve parameterized by the estimated extremal index. When the logistic map is in
a chaotic regime (e.g., `r = 3.8`) the empirical behaviour typically aligns well
with the exponential reference, lending credibility to the estimate.

