---
title: "Estimating the Extremal Index for the Logistic Map"
author: "chaoticds package"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Estimating the Extremal Index for the Logistic Map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The extremal index $\theta$ is a key parameter in extreme value theory that measures the degree of clustering in extreme events. For a stationary sequence, $\theta \in (0, 1]$ where:

- $\theta = 1$ indicates no clustering (extremes occur in isolation)
- $\theta < 1$ indicates clustering (extremes tend to occur in groups)

This vignette demonstrates how to estimate $\theta$ for the chaotic logistic map using the **chaoticds** package.

## The Logistic Map

The logistic map is defined as: $x_{n+1} = r x_n (1 - x_n)$

For $r = 3.8$, the map exhibits chaotic behavior with interesting extreme value properties.

```{r simulate}
library(chaoticds)

# Load the pre-generated dataset
data(logistic_ts)

# Basic properties
cat("Time series length:", length(logistic_ts), "\n")
cat("Range: [", round(min(logistic_ts), 3), ",", round(max(logistic_ts), 3), "]\n")
cat("Mean:", round(mean(logistic_ts), 3), "\n")
```

```{r plot-series, fig.cap="Logistic map time series (first 1000 observations)"}
plot(logistic_ts[1:1000], type = "l", 
     main = "Logistic Map Time Series (r = 3.8)",
     xlab = "Time", ylab = "Value")
```

## Threshold Selection

Choosing an appropriate threshold is crucial for extremal index estimation. We use diagnostic plots to guide our selection.

```{r threshold-selection}
# Define candidate thresholds
thresholds <- quantile(logistic_ts, probs = seq(0.85, 0.98, by = 0.01))

# Mean Residual Life plot for threshold selection
mrl_data <- mean_residual_life(logistic_ts, thresholds)
```

```{r mrl-plot, fig.cap="Mean Residual Life plot for threshold selection"}
mrl_plot(mrl_data)
```

For this analysis, we'll use the 95th percentile as our threshold:

```{r threshold}
threshold <- quantile(logistic_ts, 0.95)
cat("Selected threshold:", round(threshold, 4), "\n")

# Number of exceedances
exc <- exceedances(logistic_ts, threshold)
cat("Number of exceedances:", length(exc), "\n")
cat("Exceedance rate:", round(length(exc)/length(logistic_ts), 3), "\n")
```

## Extremal Index Estimation

### Runs Estimator

The runs estimator counts clusters based on run lengths:

```{r runs-estimator}
# Try different run lengths
run_lengths <- 1:8
theta_runs_vec <- sapply(run_lengths, function(r) {
  result <- extremal_index_runs(logistic_ts, threshold, run_length = r)
  if(length(result) == 0) NA else result
})

# Display results
results_df <- data.frame(
  run_length = run_lengths,
  theta_estimate = theta_runs_vec
)
print(results_df)
```

### Intervals Estimator

The intervals estimator provides an alternative approach:

```{r intervals-estimator}
theta_intervals <- extremal_index_intervals(logistic_ts, threshold)
cat("Intervals estimator result:", theta_intervals, "\n")
```

## Cluster Analysis

Understanding the structure of extreme clusters provides additional insight:

```{r cluster-analysis}
# Analyze cluster sizes with smaller run length to find clusters
sizes <- cluster_sizes(logistic_ts, threshold, run_length = 1)

if(length(sizes) > 0) {
  cat("Cluster statistics:\n")
  summary_stats <- cluster_summary(sizes)
  cat("  Number of clusters:", length(sizes), "\n")
  cat("  Mean cluster size:", round(summary_stats[["mean_size"]], 2), "\n")
  cat("  Cluster size variance:", round(summary_stats[["var_size"]], 2), "\n")
  cat("  Max cluster size:", max(sizes), "\n")
} else {
  cat("No clusters found\n")
}
```

## Hitting Time Analysis

Hitting times provide another perspective on extremal behavior:

```{r hitting-times}
# Calculate hitting times
hts <- hitting_times(logistic_ts, threshold)

if(length(hts) > 0) {
  cat("Hitting time statistics:\n")
  cat("  Number of hitting times:", length(hts), "\n")
  cat("  Mean hitting time:", round(mean(hts), 2), "\n")
  cat("  Median hitting time:", median(hts), "\n")
  
  # Plot hitting times if we have a valid theta estimate
  valid_theta <- theta_runs_vec[!is.na(theta_runs_vec)]
  if(length(valid_theta) > 0) {
    plot_hts(hts, valid_theta[1])
  }
}
```

## Conclusions

This analysis demonstrates the estimation of the extremal index for the chaotic logistic map. The logistic map exhibits complex extreme value behavior that can be characterized using the tools in the **chaoticds** package.

## Further Reading

- Leadbetter, M.R., Lindgren, G., & Rootzén, H. (1983). *Extremes and Related Properties of Random Sequences and Processes*
- Freitas, A.C.M., Freitas, J.M., & Todd, M. (2010). Hitting time statistics and extreme value theory. *Probability Theory and Related Fields*
- Lucarini, V. et al. (2016). *Extremes and Recurrence in Dynamical Systems*