---
title: "Getting Started with chaoticds"
author: "Diogo Ribeiro"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with chaoticds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

## Welcome!

This vignette provides a gentle introduction to the `chaoticds` package. You'll learn:

- How to simulate chaotic dynamical systems
- How to perform extreme value analysis
- How to estimate the extremal index
- How to interpret results and create visualizations

**Estimated time:** 15-20 minutes

---

## What is chaoticds?

The `chaoticds` package provides tools for analyzing **extreme events** in **chaotic dynamical systems**.

### Why is this important?

Chaotic systems appear everywhere:

- **Climate dynamics**: Temperature extremes, hurricanes
- **Financial markets**: Market crashes, volatility spikes
- **Turbulent flows**: Extreme velocities in fluids
- **Biological systems**: Population dynamics

Traditional statistical methods assume **independence**, but extreme events in chaotic systems tend to **cluster together**. The extremal index θ quantifies this clustering.

---

## Installation

```{r installation, eval = FALSE}
# From GitHub (development version)
devtools::install_github("DiogoRibeiro7/chaotic-dynamical-systems")

# From CRAN (when available)
install.packages("chaoticds")
```

Load the package:

```{r load-package}
library(chaoticds)
```

---

## Part 1: Simulating Chaotic Dynamics

### The Logistic Map

The logistic map is one of the simplest chaotic systems:

$$x_{n+1} = r \cdot x_n \cdot (1 - x_n)$$

Let's simulate it!

```{r simulate-logistic}
# Generate 1000 iterations with r = 3.8 (chaotic regime)
series <- simulate_logistic_map(n = 1000, r = 3.8, x0 = 0.2)

# Visualize
plot(series, type = "l", main = "Logistic Map Time Series (r = 3.8)",
     xlab = "Iteration", ylab = "x", col = "steelblue")
```

**What do we see?**

- No apparent pattern
- Irregular fluctuations
- This is deterministic chaos!

### Understanding the Parameter r

The dynamics change dramatically with r:

```{r parameter-exploration, fig.height = 6}
par(mfrow = c(2, 2))

# r = 2.5: Convergence to fixed point
series1 <- simulate_logistic_map(500, r = 2.5, x0 = 0.2)
plot(series1, type = "l", main = "r = 2.5 (Fixed Point)", ylab = "x", col = "darkgreen")

# r = 3.2: Period-2 oscillation
series2 <- simulate_logistic_map(500, r = 3.2, x0 = 0.2)
plot(series2, type = "l", main = "r = 3.2 (Periodic)", ylab = "x", col = "orange")

# r = 3.8: Chaos
series3 <- simulate_logistic_map(500, r = 3.8, x0 = 0.2)
plot(series3, type = "l", main = "r = 3.8 (Chaotic)", ylab = "x", col = "red")

# r = 4.0: Fully chaotic
series4 <- simulate_logistic_map(500, r = 4.0, x0 = 0.2)
plot(series4, type = "l", main = "r = 4.0 (Fully Chaotic)", ylab = "x", col = "purple")

par(mfrow = c(1, 1))
```

### Bifurcation Diagram

Visualize how dynamics change across parameter values:

```{r bifurcation}
# Generate bifurcation data
r_values <- seq(2.5, 4, length.out = 500)
bif_data <- logistic_bifurcation(r_values, n_iter = 300, discard = 250)

# Plot
plot(bif_data$r, bif_data$x, pch = ".", cex = 0.3,
     main = "Logistic Map Bifurcation Diagram",
     xlab = "Parameter r", ylab = "x",
     col = rgb(0, 0, 1, 0.3))
```

**Interpretation:**

- **r < 3**: Single values (fixed points)
- **3 < r < 3.57**: Branching pattern (period doubling)
- **r > 3.57**: Dense cloud (chaos)

---

## Part 2: Extreme Value Analysis

Now let's analyze extreme events in our chaotic time series.

### Method 1: Block Maxima

Divide the series into blocks and extract the maximum from each:

```{r block-maxima}
# Use chaotic series from r = 3.8
block_size <- 50
bm <- block_maxima(series, block_size)

cat("Original series length:", length(series), "\n")
cat("Number of block maxima:", length(bm), "\n")
cat("Block maxima range: [", round(min(bm), 3), ",", round(max(bm), 3), "]\n")

# Visualize distribution
hist(bm, breaks = 15, col = "lightblue", border = "white",
     main = "Distribution of Block Maxima",
     xlab = "Block Maximum Value", prob = TRUE)
lines(density(bm), col = "darkblue", lwd = 2)
```

### Fitting the Generalized Extreme Value (GEV) Distribution

```{r fit-gev, eval = requireNamespace("evd", quietly = TRUE)}
# Fit GEV distribution to block maxima
gev_fit <- fit_gev(bm)

# Display results
print(gev_fit)
```

**Interpretation:**

- **Location (μ)**: Center of the distribution
- **Scale (σ)**: Spread of the distribution
- **Shape (ξ)**: Tail behavior
  - ξ > 0: Heavy tail (Fréchet)
  - ξ = 0: Light tail (Gumbel)
  - ξ < 0: Bounded tail (Weibull)

### Method 2: Peaks Over Threshold (POT)

Instead of blocks, extract all values above a high threshold:

```{r pot-method}
# Select 95th percentile as threshold
threshold <- quantile(series, 0.95)
cat("Threshold:", round(threshold, 4), "\n")

# Extract exceedances
exc <- exceedances(series, threshold)
cat("Number of exceedances:", length(exc), "\n")
cat("Exceedance proportion:", length(exc) / length(series), "\n")

# Visualize
hist(exc, breaks = 10, col = "coral", border = "white",
     main = "Distribution of Exceedances",
     xlab = "Exceedance Value")
```

### Fitting the Generalized Pareto Distribution (GPD)

```{r fit-gpd, eval = requireNamespace("evd", quietly = TRUE)}
# Fit GPD to exceedances
gpd_fit <- fit_gpd(series, threshold)

# Display results
print(gpd_fit)
```

---

## Part 3: The Extremal Index

### What is the Extremal Index?

The extremal index θ ∈ (0, 1] measures clustering of extreme events:

- **θ = 1**: Extremes occur independently (like IID data)
- **θ < 1**: Extremes cluster together (typical in chaotic systems)
- **θ = 0.5**: On average, extremes come in pairs

### Estimating θ

```{r extremal-index}
# Runs estimator
run_length <- 2
theta_runs <- extremal_index_runs(series, threshold, run_length)

cat("\nExtremal Index Estimate (Runs Method):\n")
cat("θ =", round(theta_runs, 4), "\n")
cat("Interpretation:",
    ifelse(theta_runs < 0.7, "Strong clustering",
           ifelse(theta_runs < 0.9, "Moderate clustering", "Weak clustering")),
    "\n")
```

### Understanding Clusters

```{r cluster-analysis}
# Identify clusters of exceedances
sizes <- cluster_sizes(series, threshold, run_length)

cat("\nCluster Statistics:\n")
cat("Number of clusters:", length(sizes), "\n")
cat("Mean cluster size:", round(mean(sizes), 2), "\n")
cat("Max cluster size:", max(sizes), "\n")

# Visualize cluster sizes
if (length(sizes) > 0) {
  cluster_hist <- cluster_histogram(sizes)
  print(cluster_hist)
}
```

### Bootstrap Confidence Intervals

Get uncertainty estimates for θ:

```{r bootstrap-ci, eval = FALSE}
# This takes ~30 seconds
boot_result <- bootstrap_extremal_index(
  series,
  threshold,
  run_length = 2,
  B = 1000  # 1000 bootstrap samples
)

cat("Point estimate:", round(boot_result$estimate, 4), "\n")
cat("95% CI: [", round(boot_result$ci[1], 4), ",",
    round(boot_result$ci[2], 4), "]\n")
```

---

## Part 4: Threshold Selection

Choosing the right threshold is crucial! Too low → bias. Too high → high variance.

### Mean Residual Life Plot

```{r mrl-plot}
# Compute MRL for various thresholds
thresholds <- quantile(series, seq(0.85, 0.99, by = 0.01))
mrl_data <- mean_residual_life(series, thresholds)

# Create MRL plot
mrl_plot(mrl_data)
```

**How to interpret:**

- Look for linear pattern in upper tail
- Choose threshold where linearity begins
- Trade-off: higher threshold = fewer exceedances

### Hill Plot

```{r hill-plot}
# Compute Hill estimates
k_values <- 10:100
hill_data <- hill_estimates(series, k_values)

# Create Hill plot
hill_plot(hill_data)
```

**How to interpret:**

- Look for stable region (plateau)
- Estimate tail index from plateau
- More stable = better threshold choice

---

## Part 5: Diagnostic Checks

### Autocorrelation Function

Check for serial dependence:

```{r acf-check}
# Compute ACF
lags <- 1:20
acf_values <- acf_decay(series, lags)

# Plot
plot(lags, acf_values, type = "h", lwd = 6, col = "steelblue",
     main = "Autocorrelation Function",
     xlab = "Lag", ylab = "ACF",
     ylim = c(-0.2, 1))
abline(h = 0, col = "gray", lty = 2)
abline(h = c(-1.96/sqrt(length(series)), 1.96/sqrt(length(series))),
       col = "red", lty = 2)
```

### Mixing Diagnostics

Test if extremes satisfy mixing conditions:

```{r mixing}
# Check D(u_n) condition
d_result <- d_check(series, threshold, lags = 1:10)
cat("D(u_n) condition satisfied:",
    ifelse(max(abs(d_result)) < 0.1, "YES", "Possibly NO"), "\n")
```

---

## Part 6: Complete Workflow

Put it all together with `run_demo()`:

```{r complete-demo, eval = FALSE}
# Comprehensive analysis
results <- run_demo(
  n = 2000,
  r = 3.8,
  x0 = 0.2,
  block_size = 50,
  threshold_q = 0.95,
  output_report = FALSE
)

# Explore results
names(results)
#> [1] "series"          "threshold"       "diagnostics"
#> [4] "block_maxima"    "gev_fit"         "exceedances"
#> [7] "gpd_fit"         "extremal_index"  "cluster_sizes"
#> [10] "cluster_summary" "acf"             "mixing"

# Access components
summary(results$gev_fit)
print(results$extremal_index)
```

---

## Next Steps

### Learn More

- 📖 **`vignette("estimating-theta-logistic")`**: Deep dive into extremal index
- 📖 **`vignette("block-maxima-vs-pot-henon")`**: Compare EVT methods
- 📖 **`vignette("multivariate-analysis")`**: Multi-dimensional systems
- 📖 **`vignette("performance-optimization")`**: Speed up your analysis

### Try Different Systems

```{r other-systems, eval = FALSE}
# Hénon map (2D chaotic attractor)
henon <- simulate_henon_map(n = 5000, a = 1.4, b = 0.3)
plot(henon$x, henon$y, pch = ".", cex = 0.5,
     main = "Hénon Attractor")

# Tent map
tent <- simulate_tent_map(n = 1000, r = 2, x0 = 0.1)

# Lozi map
lozi <- simulate_lozi_map(n = 5000)
```

### Interactive Exploration

```{r shiny-app, eval = FALSE}
# Launch interactive Shiny app
launch_explorer()
```

---

## Getting Help

- **Function documentation**: `?simulate_logistic_map`
- **All vignettes**: `browseVignettes("chaoticds")`
- **Package website**: [https://diogoribeiro7.github.io/chaotic-dynamical-systems/](https://diogoribeiro7.github.io/chaotic-dynamical-systems/)
- **Report issues**: [GitHub Issues](https://github.com/DiogoRibeiro7/chaotic-dynamical-systems/issues)

---

## References

**Key Papers:**

1. Coles, S. (2001). *An Introduction to Statistical Modeling of Extreme Values*. Springer.

2. Freitas, A. C. M., Freitas, J. M., & Todd, M. (2010). Hitting time statistics and extreme value theory. *Probability Theory and Related Fields*, 147(3-4), 675-710.

3. Leadbetter, M. R. (1983). Extremes and local dependence in stationary sequences. *Zeitschrift für Wahrscheinlichkeitstheorie und verwandte Gebiete*, 65(2), 291-306.

**Further Reading:**

- Embrechts, P., Klüppelberg, C., & Mikosch, T. (1997). *Modelling Extremal Events*. Springer.

---

## Summary

You've learned:

✅ How to simulate chaotic dynamical systems
✅ How to perform extreme value analysis (block maxima & POT)
✅ How to estimate and interpret the extremal index
✅ How to select thresholds using diagnostic plots
✅ How to check model assumptions

**Happy analyzing! 🎉**
