---
title: "chaoticds: end-to-end walkthrough"
author: "Diogo Ribeiro"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE
)
set.seed(20260513)
library(chaoticds)
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
```

A compact tour of the package: **simulate → block maxima / POT → extremal index →
plot**, in one short narrative. For deeper treatment of any step, see the
[package vignettes](https://diogoribeiro7.github.io/chaotic-dynamical-systems/articles/).

## 1. Simulate a chaotic orbit

We generate a logistic-map trajectory in the fully chaotic regime (`r = 3.8`).
A short transient is discarded by `simulate_logistic_map()` so the orbit lives
on the attractor from step 1.

```{r simulate}
n       <- 5000
series  <- simulate_logistic_map(n = n, r = 3.8, x0 = 0.2)

ggplot(data.frame(t = seq_len(n), x = series), aes(t, x)) +
  geom_line(colour = "steelblue", linewidth = 0.3) +
  labs(title = "Logistic map (r = 3.8)", x = "iteration", y = expression(x[t]))
```

## 2. Block maxima → GEV

Split the series into blocks of size 50 and fit a Generalised Extreme Value
distribution to the per-block maxima.

```{r block-maxima}
bm <- block_maxima(series, block_size = 50)

ggplot(data.frame(bm), aes(bm)) +
  geom_histogram(bins = 25, fill = "steelblue", colour = "white") +
  labs(title = sprintf("Block maxima (n = %d, block size = 50)", length(bm)),
       x = "block maximum", y = "count")
```

```{r fit-gev, eval = requireNamespace("evd", quietly = TRUE)}
gev <- fit_gev(bm)
print(gev)
```

## 3. Peaks over threshold → GPD

Same data, different lens: take all values above the 95th percentile and fit a
Generalised Pareto distribution to the excesses.

```{r pot}
threshold <- quantile(series, 0.95)
exc       <- exceedances(series, threshold)

ggplot(data.frame(t = which(series > threshold), x = series[series > threshold]),
       aes(t, x)) +
  geom_segment(aes(xend = t, yend = threshold), colour = "firebrick", alpha = 0.6) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  labs(title = sprintf("Exceedances over u = %.3f (%d points)",
                       threshold, length(exc)),
       x = "iteration", y = expression(x[t]))
```

```{r fit-gpd, eval = requireNamespace("evd", quietly = TRUE)}
gpd <- fit_gpd(series, threshold)
print(gpd)
```

## 4. Extremal index — do extremes cluster?

For an IID sequence θ = 1. Chaotic dynamics typically give θ < 1, meaning
extremes arrive in clusters. Two estimators agree to within bootstrap noise:

```{r theta}
theta_runs      <- extremal_index_runs(series, threshold, run_length = 2)
theta_intervals <- extremal_index_intervals(series, threshold)

c(runs = theta_runs, intervals = theta_intervals)
```

A block-bootstrap confidence interval quantifies the uncertainty (small `B` here
to keep the walkthrough fast — bump to 1000+ for real work):

```{r theta-bootstrap}
boot <- bootstrap_extremal_index(
  series, threshold,
  estimator   = "runs",
  run_length  = 2,
  B           = 200,
  seed        = 1
)

sprintf("theta = %.3f   95%% CI = [%.3f, %.3f]",
        boot$estimate, boot$ci[1], boot$ci[2])
```

## 5. Cluster structure

Group consecutive exceedances into clusters and inspect their sizes — the
empirical mean cluster size is roughly `1/θ`.

```{r clusters}
sizes <- cluster_sizes(series, threshold, run_length = 2)
summary(sizes)
cat(sprintf("1/theta_runs = %.2f, mean(cluster size) = %.2f\n",
            1 / theta_runs, mean(sizes)))
```

```{r cluster-plot, fig.height = 3}
ggplot(data.frame(size = sizes), aes(factor(size))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Cluster size distribution", x = "cluster size", y = "count")
```

## 6. One-shot: `run_demo()`

Everything above is bundled into `run_demo()`, which is convenient when you
want the full pipeline as a named list (handy for reports or unit tests):

```{r run-demo, eval = FALSE}
results <- run_demo(n = 2000, r = 3.8, x0 = 0.2,
                    block_size = 50, threshold_q = 0.95,
                    output_report = FALSE)
str(results, max.level = 1)
```

## Where to next

- **Theory & deeper examples** — start with the `getting-started` vignette, then
  `estimating-theta-logistic` and `block-maxima-vs-pot-henon`.
- **Higher-dimensional dynamics** — see `multivariate-analysis`.
- **Performance** — for series with >1e6 points, see the `*_cpp` / `*_smart`
  variants and the `performance-optimization` vignette.
- **Interactive** — `launch_explorer()` opens a Shiny app for hands-on
  parameter exploration.
